---
title: "meditation-tracking"
author: "Matti Heino"
date: "26 joulukuuta 2017"
output: html_document
---

Set up paths and packages.

```{r setup, include=FALSE}

if (!require(pacman)) install.packages("pacman"); library(pacman)

Sys.setenv(R_MAX_NUM_DLLS = 1000)
Sys.getenv("R_MAX_NUM_DLLS")


p_load(tidyverse, devtools, rio, plyr, tidyverse, Matrix, ggplot2, 
       lattice, latticeExtra, grid, gridExtra, scales, dygraphs,
       rgl, plot3D, fractal, nonlinearTseries, crqa, signal, sapa, 
       ifultools, pracma, nlme, lme4, lmerTest, minpack.lm, igraph,
       qgraph, graphicalVAR, IsingSampler, IsingFit, lubridate,
       randtests, TSA, tseries, forecast, tidyverse, lubridate, zoo,
       userfriendlyscience, dummies) # bootGraph not found

p_load_gh("AtherEnergy/ggTimeSeries")

knitr::opts_chunk$set(echo = TRUE, 
               warning = TRUE,
               error = TRUE,
               cache = TRUE, 
               collapse = TRUE,
               eval = TRUE)
knitr::opts_chunk$set(root.dir = ".")  # Always project root as working directory
knitr::opts_knit$set(root.dir = ".")  # This is needed for some versions of RStudio

```

Set up data.

The data file is the worksheet "Kopio: Sheet1" exported from Google Sheets to Excel and then to csv. If file is directly exported to csv, it uses commas as separators but keeps commas as decimal. Going through Excel makes the separator ";", which is properly read by read_csv2.  

```{r}

data <- read_csv2("data/Headspace v2 1-50, 207-236, 319-357 (2).csv")

# Fix the comma as decimal point in the variable r1.
data$r1 <- sub(",", data$r1, replacement = ".") %>% as.numeric(.)

# There were 5 congruent trials and 15 incongruent ones each time.
# The stroop test program divided the total time for congruent trials by 5
# to get the average but, for incongruent trials, by 10 instead of 15.
# Rows 2 and 5 in code below fix this.

data <- data %>% dplyr::mutate(
  cong_pre = kongaika1,
  incong_pre = inkongaika1 * 15 / 10,
  diff_pre = incong_pre - cong_pre,
  cong_post = kongaika2,
  incong_post = inkongaika2 * 15 / 10,
  diff_post = incong_post - cong_post,
  diff_postpre = diff_post - diff_pre,
  selfRegulationFailure = `POV (0=ei, 1=puuhat, 2=palvonta tms., 3=mind, 4=POV)`) %>% 
  dplyr::filter(as == "1" | as == "2")

# Divide this variable to dummy categories
dummy.data.frame(data, names = data$selfRegulationFailure)
selfRegulationFailure <- dummy(data$selfRegulationFailure) %>%
  data.frame

names(selfRegulationFailure) <- c("selfRegulationFailure0", "selfRegulationFailure1", "selfRegulationFailure2", "selfRegulationFailure3", "selfRegulationFailure4", "selfRegulationFailureNA")

# Stitch the dummies together with main data
data <- bind_cols(data, selfRegulationFailure) 

# NA is now zero on the dummies. Fix it.
data <- data %>% rowwise %>% 
  mutate(selfRegulationFailure0 = ifelse(selfRegulationFailureNA == 1, NA, selfRegulationFailure0),
  selfRegulationFailure1 = ifelse(selfRegulationFailureNA == 1, NA, selfRegulationFailure1),
  selfRegulationFailure2 = ifelse(selfRegulationFailureNA == 1, NA, selfRegulationFailure2),
  selfRegulationFailure3 = ifelse(selfRegulationFailureNA == 1, NA, selfRegulationFailure3),
  selfRegulationFailure4 = ifelse(selfRegulationFailureNA == 1, NA, selfRegulationFailure4))

# temp: trying different variables
data$diff_postpre <- data$v1                               

# This creates the date and time; a lot of NAs
# data$time <- dmy_hm(paste0(data$pva_korj, " ", data$klo), tz = "Europe/Helsinki")

# This creates just the date
data$time <- dmy(data$pva_eikorj, tz = "Europe/Helsinki")

```

```{r}
data2 <- data.frame(data$s1, data$r1, data$v1)
associationMatrix(data2)

associationMatrix(data.frame(data$v1, data$r1, data$s1, data$selfRegulationFailure4))
cor(data.frame(data$v1, data$r1, data$s1, data$selfRegulationFailure4), use = "pairwise.complete.obs", method = "kendall")

```

```{r}
data %>% dplyr::filter(no >= 500) %>% 
  ggplot(aes(x = no, y = diff_postpre), group = 1) +
  geom_line() +
  geom_point(aes(x = no, y = as.numeric(selfRegulationFailure)), colour = "darkblue", size = 1)


```

# Tests for level and trend stationarity 

## Bartels rank test
```{r}

randtests::bartels.rank.test(data$diff_postpre, alternative = "two.sided")
randtests::bartels.rank.test(data$diff_postpre, alternative = "left.sided")
randtests::bartels.rank.test(data$diff_postpre, alternative = "right.sided")

```

The results indicate that the data may be non-random, with a trend, without systematic oscillation.

## Cox-Stuart sign test 

```{r}

randtests::cox.stuart.test(na.exclude(data$diff_postpre), alternative = "two.sided")
randtests::cox.stuart.test(na.exclude(data$diff_postpre), alternative = "left.sided")
randtests::cox.stuart.test(na.exclude(data$diff_postpre), alternative = "right.sided")

```

The Cox-Stuart test indicates no upward nor downward trend, and random values.

## KPSS test 

```{r}

tseries::kpss.test(na.exclude(data$diff_postpre), lshort = TRUE, null = "Level")
tseries::kpss.test(na.exclude(data$diff_postpre), lshort = TRUE, null = "Trend")

```

## Test for AR, ARCH or an optimal ARIMA process

First inspect the partial autocorrelation function to get an idea of the AR order.

```{r}
par(mfrow=c(1,2))
acf(data$diff_postpre, na.action = na.pass)
pacf(data$diff_postpre, na.action = na.pass)
```

```{r}
# KEENAN 1-DEGREE TEST OF NONLINEARITY
# H0: time series follows some AR process
# H1: time series cannot be considered some AR process

TSA::Keenan.test(na.exclude(data$diff_postpre))

```

We cannot rule out that this data follows some AR process.

```{r}
# MCLEOD-LI TEST FOR CONDITIONAL HETEROSCEDASTICITY
# H0: time series follows some AR process
# H1: time series cannot be considered some ARCH process

TSA::McLeod.Li.test(y = data$diff_postpre, plot = TRUE, omit.initial = TRUE)
```

We cannot rule out that this data follows some AR process.

```{r}
# MCLEOD-LI TEST FOR CONDITIONAL HETEROSCEDASTICITY
# H0: time series follows some AR process
# H1: time series cannot be considered some ARiMA process

TSA::McLeod.Li.test(object = forecast::auto.arima(data$diff_postpre), plot = TRUE, omit.initial = TRUE)
```

```{r}
beepMean <- mean(time_length(diff(data$time), unit = "hours"), na.rm = TRUE)

ts0 <- data$diff_postpre[3:length(data$diff_postpre)] %>% ts(., frequency = round(beepMean))
fit <- StructTS(ts0, type="BSM")
par(mfrow = c(4, 1)) # to give appropriate aspect ratio for next plot.
plot(cbind(fitted(fit), resids=resid(fit)), main= "Basic Structural Model for 'diff_postpre'")
```

